% BEGIN LICENSE BLOCK
% Version: CMPL 1.1
%
% The contents of this file are subject to the Cisco-style Mozilla Public
% License Version 1.1 (the "License"); you may not use this file except
% in compliance with the License.  You may obtain a copy of the License
% at www.eclipse-clp.org/license.
% 
% Software distributed under the License is distributed on an "AS IS"
% basis, WITHOUT WARRANTY OF ANY KIND, either express or implied.  See
% the License for the specific language governing rights and limitations
% under the License. 
% 
% The Original Code is  The ECLiPSe Constraint Logic Programming System. 
% The Initial Developer of the Original Code is  Cisco Systems, Inc. 
% Portions created by the Initial Developer are
% Copyright (C) 2006 Cisco Systems, Inc.  All Rights Reserved.
% 
% Contributor(s): 
% 
% END LICENSE BLOCK


% File		: manual.tex
% Date		: March 1992
% Author	: Michael Dahmen
% Modified by	: Luis Hermosilla, August 1992
% Project	: MegaLog-Sepia User Manual
% Content	: Root file

%
%		Preamble
%

\documentstyle[11pt,html,a4wide,epsf]{book}
\include{sepiachip}

%
% Document
%

\begin{document}

\bibliographystyle{plain}

\title{
    {\Large \eclipse}\\
    \vspace{1cm}
    {\Huge Knowledge Base User Manual}\\
    \vspace{1cm}
    Release 3.4}
\date{July 1994}
\author{ }

\maketitle

% Needed to adjust left/right pages properly
\setcounter{page}{2}
% Suppress printing of the page number on this page
\pagestyle{empty}


Copyright \copyright\ 1992 -- 2006 Cisco Systems, Inc.

\cleardoublepage
\pagestyle{plain}
\pagenumbering{roman}

\begin{latexonly}
\tableofcontents
\end{latexonly}

\cleardoublepage
\pagenumbering{arabic}

\include {intro-sec}
% What is MegaLog-Sepia

\include {database-sec}
% Database connection 

\include {knowbase-sec}
% Deductive Database connection

\include {multiuser-sec}
% Features to support multi-user MegaLog

\include {backwards-sec}
% compatibility with previous MegaLog

% \include {profile-sec}
% Profiler manual

\chapter{The Built-In Predicates}
\label{bip-summary}

This chapter provides a summary of each of the built-in 
predicates available in the database and knowledge base of \eclipse. The
built-ins are listed 
in alphabetical order within each type of built-in. The information
provided here is only a one-line description for each built-in.
A more complete specification can be found by
referring to the Knowledge Base Built-In Specification Manual (or Knowledge
Base BIP Book), 
\cite{BIP92}. The Knowledge Base BIP Book is on-line available by the
predicate {\bf help/1}.

{\bf help(Functor/Arity)} prints the page from the Knowledge Base BIP Book
describing the  
built-in {\bf Functor/Arity} to the current output. If only {\bf Functor}
is specified, the system looks for all predicates that have this atom in 
their name. If there are more than one predicate
matching, a short description for each one is printed, rather than the
whole page. There is also a glossary, where each page does not describe
a single built-in but a concept that is used in other descriptions.
Examples :

\begin{verbatim}
% display page from Knowledge Base BIP Book
?- help(createdb/1).

% list a short description of all built-ins matching "relation"
?- help(relation).    

% list a short description of all glossary entries
?- help(glossary).

% display glossary on Attribute Specification
?- help(attribute_specification).
\end{verbatim}

The built-ins are classified into the following types,
which correspond to the defining module.

\paragraph{List of built-in types:}

\begin{enumerate}

\item{\eclipse Database Kernel}
\item{\eclipse DB}
\item{\eclipse KB}

\end{enumerate}

\include {kernel-lst}
% database predicates

\include {database-lst}
% relational algebra predicates

\include {knowbase-lst}
% knowledge base predicates

\newpage
\include {indexfile}
\newpage

\bibliography{sepiachip}

\end{document}
