% BEGIN LICENSE BLOCK
% Version: CMPL 1.1
%
% The contents of this file are subject to the Cisco-style Mozilla Public
% License Version 1.1 (the "License"); you may not use this file except
% in compliance with the License.  You may obtain a copy of the License
% at www.eclipse-clp.org/license.
% 
% Software distributed under the License is distributed on an "AS IS"
% basis, WITHOUT WARRANTY OF ANY KIND, either express or implied.  See
% the License for the specific language governing rights and limitations
% under the License. 
% 
% The Original Code is  The ECLiPSe Constraint Logic Programming System. 
% The Initial Developer of the Original Code is  Cisco Systems, Inc. 
% Portions created by the Initial Developer are
% Copyright (C) 2006 Cisco Systems, Inc.  All Rights Reserved.
% 
% Contributor(s): 
% 
% END LICENSE BLOCK

\chapter{DBI: \eclipse\ SQL Database Interface}
\label{chapdbi}
%HEVEA\cutdef[1]{section}
\index{database, interface to|(}
\index{MySQL, interface to|(}
%--------------------------------------------------------------------
\section{Introduction}
%--------------------------------------------------------------------

The \eclipse\ SQL database interface is a low-level interface to the SQL
language. As far as possible it has been attempted to give the
full power of the SQL interface to the programmer.

A number of features are designed to permit transfer of large amounts
of data with minimal overhead:
\begin{itemize}
\item The buffered retrieval of multiple tuples from
a relation or view.
\item Buffered multiple updates and inserts
\item Re-use of SQL statements that are parametrised with placeholders.
\item Tuple templates to permit the allocation  and re-use of fixed size
buffers.
\end{itemize}

The interface provides safe handles to database cursors. Higher
level interfaces can be constructed on top of this. For
example:

\begin{itemize}
\item Viewing an SQL query as a predicate that yields
different tuples on backtracking.
\item Viewing an SQL query as a lazily constructed list of tuples.
\end{itemize}

On backtracking, cursors as well as database sessions are automatically
closed. This is required  if one is to build abstractions where they
are hidden from the programmer.

Nowhere in the interface are SQL commands ever looked
into. They are always passed straight from the user to the database.
This means that any SQL command supported by the underlying database will
be accessible. Note also any differences in the SQL between different databases
are also not hidden from the user, so the SQL written by the user must be
valid SQL for the database that is being interfaced to.

When retrieving or inserting tuples, tuple templates are used to
specify the types of the different fields being retrieved or
inserted. This allows for a flexible mapping between database
internal types and the tuples retrieved. For example one can
retrieve a date as either a string or an atom. What type combinations
are supported will depend on the underlying database.

It is possible to write SQL queries with parameters (SQL placeholders),
so that one can use the same query several times, with different
values. This feature, together with the availability of any number
of cursors can be used to write complex queries not
expressible with a singe SQL statement.
 
If the database has a generic binary format, such as Binary Long Object
(BLOB) fields, one can
store and retrieve arbitrary terms in them automatically using the
interface. To exchange
{\eclipse\/} terms with other applications via the database, the terms can
be converted to EXDR data interchange format (see the Embedding and
Interfacing chapter for details) before storing them in the database.

It is possible to open several sessions to different databases
simultaneously. Transactions apply to one session only though.

The code is written so that it will be relatively easy to extend it to
interface to different kinds of databases. The currently supported database is
MySQL.

%--------------------------------------------------------------------
\section{Using the SQL database interface}
%--------------------------------------------------------------------
The SQL database interface is contained in the dbi library.
\begin{verbatim}
[eclipse 1]: lib(dbi). 
...
yes.
\end{verbatim}

Your environment must be set up so that you can connect to a
database supported by lib(dbi). Normally a database administrator will 
have written a script to do this automatically. 

MySQL specific note: on some platforms, the MySQL client library is
provided as a dynamic load library (libmysqlclient.so in most Unix systems,
libmysql.dll in Windows systems). This file is
not provided with {\eclipse} distribution as it is part of the MySQL
system, and is covered by its own license. MySQL can be downloaded from 
\ahrefurl{http://dev.mysql.com/downloads}.
To successfully load the dbi library in such cases, you must have a version
of this library file that match the one that your copy of {\eclipse} was
compiled with, and in addition, {\eclipse} must be able to find this
library file when the dbi library is loaded.  Normally, the person who
installed your {\eclipse} system should also make sure that {\eclipse} can
find this client library file using the normal way that the operating
system can find dynamic load libraries, e.g.\ by putting the file in a
standard system library location, or in the {\eclipse} dynamic library
directory for the platform in your {\eclipse} directory. On Unix systems,
it is also possible for the user to set an environment variable (usually
{\tt LD_LIBRARY_PATH}) to point to where the dynamic library is.
 
%--------------------------------------------------------------------
\section{Data Templates}
%--------------------------------------------------------------------
\label{data-templates}

If supported by the database, the interface allows the use of
prepared SQL statements with parameters (placeholders). Prepared SQL
statements are pre-parsed by the database, and can be executed more
efficiently for multiple times, with the placeholders acting like variables,
taking on different values for each execution.

The syntax used for
prepared statements is that provided by the database, but a common syntax
is to use \verb'?' to indicate a placeholder. For example:
\begin{verbatim}
insert into employees (enum, ename, esalary, ejob) values (?, ?, ?, ?)
\end{verbatim}
would be used to add rows to the employees relation.

Such an SQL statement has to be prepared before execution. It can then
be executed in batches to insert several tuples in one go. Preparation
involves parsing the SQL statement and setting up a buffer for the tuples.

A data template is used as an example buffer. For the insert command above it
might look like:

\begin{verbatim}
emp(1234,"some name",1000.0,'some job')
\end{verbatim}

The argument positions correspond to the order of the placeholder in the
SQL statement.
The types of the data will be used to type-check the tuples when they
are inserted.

The following \eclipse\ goal uses a template to create a cursor for an insert
command:
\begin{verbatim}
SQL = "insert into employees (enum,ename,esalary,ejob) values (?,?,?,?)",
Template = emp(1234,"some name",1000.0,'some job'),
session_sql_prepare(H, Template, SQL, Cursor),
\end{verbatim}
\verb'H' is a handle to a database session, and \verb'Cursor' is the cursor
created for the prepared statement \verb'SQL'.

The cursor can then be used to insert several rows into the employee table.

\begin{verbatim}
cursor_next_execute(Cursor,emp(1001,"E.G. SMITH",1499.08,editor)),
cursor_next_execute(Cursor,emp(1002,"D.E. JONES",1499.08,journalist)),
\end{verbatim}

Similarly for queries a data template specifies the types of the
columns retrieved. The positions of the arguments correspond to
the position of the select list items. The example template above
might be used for a query like

\begin{verbatim}
SQL = "select enum, ename, esalary, ejob from employees where esalary > 1000",
Template = emp(1234,"some name",1000.0,'some job'),
session_sql_query(H, Template, SQL, Cursor),
cursor_next_tuple(Cursor,Tuple),
% Tuple is now somthing like emp(1001,"E.G. SMITH",1499.08,editor)
\end{verbatim}

If a structure or list appears in
one of the argument positions this stands for a general term, to
be stored or retrieved in external database format. This way one is not
limited to atomic types which have natural mappings to database types.

%--------------------------------------------------------------------
\subsection{Conversion between \eclipse\ and database types}
%--------------------------------------------------------------------

Data is passed from {\eclipse\/} into the database via placeholders in
prepared SQL statements, and passed from the database to {\eclipse\/} via
the results tuples returned by executing SQL queries. 
The interface takes care of the conversion of data to/from {\eclipse\/}
types to the external C API types, and the database then converts these to/from
the internal types of the database, as determined by the SQL statement used. 
The exact internal database types, and the exact conversion rules between
the C type and the database type is dependent on the database's API, but in
general the following should hold for {\eclipse\/} types:

\begin{description}
\item[Strings and atoms] are converted to C char * type. This should be used for 
  non-numeric data. Restrictions may apply depending on the SQL datatype --
  for example, non-binary string types (such as VARCHAR) does not accept
  generic binary strings, and SQL data and time types must be in the
  correct syntax -- consult the database manual for the syntax for these types.
\item[Integers and Floats] are converted to C integers and doubles, which are
  then converted to the specified SQL numeric types. The numbers are passed
  to the database's C API at the maximum precision and size supported by
  the database's API. Any integers outside the range representable by the
   C API's integer type will raise an error. Note that while the number
  passed to the database is at maximum precisiion and size, the
  corresponding SQL numberic type specified by the SQL statement that
  receives the number may be smaller
  (e.g. SMALLINT). The exact behaviour in this case
  depends on the database.
\item[General terms] are converted to {\eclipse}'s internal dbformat
 - a flat binary representation of the term, and then to an appropriate SQL
 binary type. This 
 allows {\eclipse\/} to store and retrieve general terms, but if it is
 required to exchange Prolog terms with external sources via the
 database, then the term should be first converted to EXDR format, and the
 EXDR string can then be passed to the database. Note that EXDR can only
 represent a subset of terms -- see the Embedding and Interfacing manual
 for details.

\end{description}


%--------------------------------------------------------------------
\subsection{Specifying buffer sizes in templates}
%--------------------------------------------------------------------
Prolog terms, strings and atoms can have variable sizes, but when they are
passed into and out of the database, they pass through fixed size buffers
for reasons of efficiency.

In the case of fetching data from fixed size database fields, the size of
these buffers is by default, the size of the field in the database.
In the case of variable sized fields and of placeholders, a default size
is chosen.

Rather than taking the default it is possible to write templates that
specify a size. This is done by mentioning the size in the argument of
the template.

\begin{description}
\item['123'] defines an atom datatype where the maximum length in
bytes is 123. The length is given as a decimal number.

\item["123"] defines a string datatype where the maximum length in
bytes is 123. The length is given as a decimal number.

\item[s(123,X)] Describes any Prolog term that occupies up to 123 bytes
in external database format. Any structure whose first argument is an integer
can be used.
\end{description}

For example the following two templates specify the same type of tuple
but the second one defines some sizes for the different elements in the
tuple as well.
\begin{verbatim}
emp(1234,"name", Rules, job)
emp(1234,"10",rules(4000),'10')
\end{verbatim}

The size information is used to define the minimum size of the buffer used
to pass data between ECLiPSe and the database. Depending on the database
and the situation (input/output, prepared/direct statements), such an 
intermediate buffer may not be used; in such cases, the buffer size will be
ignored. Otherwise, if the data is too big to fit into the buffer, an error
will be raised. 

The data in the buffer is passed to/from the database, which may have its own
size specification for the data, which is independent of the size
information specified in the template. In the case of sending data to the
database, and the data is too large for the database, the exact behaviour
is dependent on the database. In the case of receiving the data from the
database, and the data is too large for the buffer, an error will be raised.

%--------------------------------------------------------------------
\section{Built-Ins}
%--------------------------------------------------------------------

%--------------------------------------------------------------------
\subsection{Sessions}
%--------------------------------------------------------------------

These predicates deal with sessions as a whole. A session is 
used to identify a connection to a database. Associated to that
session are a number of cursors. These are handles to SQL statements
which are currently being executed. For example a query where only
some of the matching rows have been fetched.


Database transactions -- where updates to the database are local to the
session until committed, and where uncommitted changes can be rolled back, 
are supported if the external database supports transactions\footnote{Some
databases supports both transactional and non-transactional updates,
and not all updates can be rolled back. Consult the database manual for
more details}. 

\subsubsection{\biptxtref{session_start(+Login, +Password, +Options, -Session)}{session_start/4}{../bips/lib/dbi/session_start-4.html}}
\label{session-start/3}
%--------------------------------------------------------------------

This create a new session by establishing a new connections to the
database server, and associates it with a handle,
returned in Session.

The session remains in existence in subsequent goals. It is automatically
closed if the program fails or aborts back beyond this call. The session
is used as a access handle to the database in subsequent calls to this
interface.

The automatic closure is particularly useful in case of a
program aborting due to a runtime error. Closing the database ensures
any database updates that have not been committed will be undone.


\subsubsection{\biptxtref{session_close(+Session)}{session_close/1}{../bips/lib/dbi/session_close-1.html}}
\label{session-close/1}
%--------------------------------------------------------------------

This closes a session, disconnecting from the database server. It
takes effect immediately. This allow resources allocated for the
session to be freed. To free all resources associated with a session,
all cursors of the session should also be closed with
\bipref{cursor_close/1}{../bips/lib/dbi/cursor_close-1.html}. 


\subsubsection{\biptxtref{session_commit(+Session)}{session_commit/1}{../bips/lib/dbi/session_commit-1.html}}
\label{session-commit/1}
%--------------------------------------------------------------------

If executed outside the scope of a
\bipref{session_transaction/2}{../bips/lib/dbi/session_transaction-2.html}
goal, this commits any transactional updates to the database made within
 Session. 
 Otherwise, it simply succeeds without doing anything.

\subsubsection{\biptxtref{session_rollback(+Session)}{session_rollback/1}{../bips/lib/dbi/session_rollback-1.html}}
\label{session-rollback/1}
%--------------------------------------------------------------------

If executed outside the scope of a
\bipref{session_transaction/2}{../bips/lib/dbi/session_transaction-2.html}
goal,
this undoes all transactional changes made to the database since the last
commit for this session. Otherwise, it will simply abort the complete outer
transaction. (Note: not all changes can be rolled back; consult the DB manual for details)

\subsubsection{\biptxtref{session_transaction(+Session, +Goal)}{session_transaction/2}{../bips/lib/dbi/session_transaction-2.html}}
\label{session-transaction/2}
%--------------------------------------------------------------------

This executes Goal as a database transaction. This predicate is only useful
if the database supports transactions. Data base updates within Goal are
committed if Goal succeeds; if Goal fails or aborts, the updates are rolled
back. 

Calls of this predicate can be nested, but only the outermost
transaction is real. All the inner transactions are simply
equivalent to call(Goal).
This way it is possible to write a call to
session_transaction/2, into some code that implements a simple
update, but then to include that simple update into a larger
transaction.

Transactions are local to one session so there is no way to safely
make an update relating to several sessions.

\begin{verbatim}
recorded_transfer(Session,Date,Amount,FromAccount,ToAccount) :-
    session_transaction(Session, (
        transfer(Session, Amount,FromAccount,ToAccount),
        check_overdraft_limit(FromAccount),
        record_transfer(Date,Amount,FromAccount,ToAccount)
    )).

transfer(Session, Amount,FromAccount,ToAccount) :-
     session_transaction(Session, 
         transfer_(Session,Amount,FromAccount,ToAccount)
     ).

\end{verbatim}

In the above example we can see two nested transactions. One simple bank
transfer that is not recorded, and an outer transaction recording the
occurrence of the transfer and checking the balance.

Since a nested transaction is simply a call of its goal, with no
partial rollbacks care has to be taken not to redo transactions on
failure unless one is sure one is at an outer transaction.

%--------------------------------------------------------------------
\subsection{Database Updates}
%--------------------------------------------------------------------


For database updates, lib(dbi) provides predicates to execute SQL
statements on the database without returning results. 
\bipref{session_sql/3}{../bips/lib/dbi/session_sql-3.html} executes
an SQL statement directly. \bipref{session_sql_prepare/4}{../bips/lib/dbi/session_sql_prepare-4.html} is used to
prepare SQL statements, returning a cursor to the prepared statement,
which can then be executed
multiple times with different placeholder values using either
\bipref{cursor_next_execute/2}{../bips/lib/dbi/cursor_next_execute-2.html} or
\bipref{cursor_all_execute/2}{../bips/lib/dbi/cursor_all_execute-2.html} or
\bipref{cursor_N_execute/4}{../bips/lib/dbi/cursor_N_execute-4.html}.
Cursors are automatically closed if the program backtracks or aborts beyond
the predicate that created it. Alternatively, the cursor can be closed
explicitly by \bipref{cursor_close/1}{../bips/lib/dbi/cursor_close-1.html}.

The datatypes of the parameters for the prepared statement is specified by
a template given to session_sql_prepare/4. See section~\ref{data-templates}
for details on the templates.  

\subsubsection{\biptxtref{session_sql(+Session, +SQL, -RowProcessedCount)}{session_sql/3}{../bips/lib/dbi/session_sql-3.html}}
\label{session-sql/3}
%--------------------------------------------------------------------

This is the simplest interface to execute an SQL statement with no
placeholders.

\begin{verbatim}
make_accounts(Session) :-
    session_sql(Session,
        "create table accounts \
         (id           decimal(4)      not null,\
          balance      decimal(9,2)    default 0.0 not null, \
          overdraft    decimal(9,2)    default 0.0 not null \
         )" ,_),
    session_sql(Session,
        "insert into accounts (id,balance) values (1001,1200.0)",1),
    session_sql(Session,
        "insert into accounts (id,balance) values (1002,4300.0)",1).
\end{verbatim}
In the example we see session_sql/3 used, first to create a
table, and then to initialise it with two rows. The rows processed counts
are checked to make sure exactly one row is processed per statement.

This code assumes a session with handle {\tt Handle} has been started beforehand.

\subsubsection{\biptxtref{session_sql_prepare(+Session, +Template, +SQL, -Cursor)}{session_sql_prepare/4}{../bips/lib/dbi/session_sql_prepare-4.html}}
\label{session-sql-prepare/4}
%--------------------------------------------------------------------

This creates Cursor, which is a handle to the prepared statement,
possibly with placeholders. Template specifies the types of the placeholders 
(see section~\ref{data-templates}). 

\begin{verbatim}
transfer_(Session, Amount, FromAccount, ToAccount) :-
    SQL = "update accounts set balance = balance + ? \
                                             where id = ?",
    Deduct is - Amount,
    session_sql_prepare(Session,incbal(1.0,12),SQL,1,Update),
    cursor_next_execute(Update,incbal(Deduct,FromAccount)),
    cursor_next_execute(Update,incbal(Amount,ToAccount)).
\end{verbatim}
In the example a cursor is prepared to modify account balances. It is used
twice, once to deduct an amount and once to add that same amount to
another account. Note: the example uses MySQL's syntax for prepared
statement, which may differ from other databases. Consult your database manual for
prepared statement syntax.

\subsubsection{\biptxtref{cursor_next_execute(+Cursor, +Tuple)}{cursor_next_execute/2}{../bips/lib/dbi/cursor_next_execute-2.html}}
\label{cursor-next-execute/2}
%--------------------------------------------------------------------

Execute the prepared SQL statement represented by Cursor, with Tuple
supplying the values for any parameter values.
This call can be executed any number of times on the same cursor.

\subsubsection
{\biptxtref{cursor_all_execute(+Cursor, +TupleList)}{cursor_all_execute/2}{../bips/lib/dbi/cursor_all_execute-2.html}}
\label{cursor-all-execute/2}
%--------------------------------------------------------------------

The SQL statement of Cursor is executed once for each Tuple in 
TupleList. This could be defined as:
\begin{verbatim}
cursor_all_execute( Cursor, []).
cursor_all_execute( Cursor, [Tuple | Tuples] ) :-
    cursor_next_execute(Cursor, Tuple),
    cursor_all_execute( Cursor, Tuples ).
\end{verbatim}

\subsubsection
{\biptxtref{cursor_N_execute(+Cursor, -Executed, +TupleList, -RestTupleList)}{cursor_N_execute/4}{../bips/lib/dbi/cursor_N_execute-4.html}}
\label{cursor-N-execute/4}
%--------------------------------------------------------------------

Some databases supports the execution of multiple tuples of parameter values at
once, doing this more efficiently than executing each tuple of parameter
values one by one. This predicate is provided to support this. 

Note that for databases that does not support execution of multiple tuples, this
predicate is implemented by executing the Tuples one by one as in 
\bipref{cursor_next_execute/2}{../bips/lib/dbi/cursor_next_execute-2.html},
and there is no gain in efficiency over using cursor_next_execute/2.

\begin{verbatim}
transfer_(Session, Amount, FromAccount, ToAccount) :-
    SQL = "update accounts set balance = balance + ? \
                                             where id = ?",
    Deduct is - Amount,
    session_sql_prepare(Session,incbal(1.0,12),SQL,2,Update),
    Updates = [incbal(Deduct,FromAccount),incbal(Amount,ToAccount)],
    cursor_N_execute(Update,2,Updates,[]).
\end{verbatim}

The example shows how to re-code the bank transfer predicate from
\bipref{cursor_next_execute/2}{../bips/lib/dbi/cursor_next_execute-2.html},
to execute
both updates with one call. This could lead to some performance
improvement in a client server setting for databases that supports multiple
parameter tuples.

%--------------------------------------------------------------------
\subsection{Database Queries}
%--------------------------------------------------------------------

For database queries, lib(dbi) provides predicates to execute SQL
statements and extract the results returned by the database.
\bipref{session_sql_query/4}{../bips/lib/dbi/session_sql_query-4.html} or
\bipref{session_sql_query/5}{../bips/lib/dbi/session_sql_query-5.html} 
executes an SQL statement, returning a cursor to allow the results to be
extracted from the database. The predicates to do this are
\bipref{cursor_next_tuple/2}{../bips/lib/dbi/cursor_next_tuple-2.html},
\bipref{cursor_all_tuples/2}{../bips/lib/dbi/cursor_all_tuples-2.html} and
\bipref{cursor_N_tuples/4}{../bips/lib/dbi/cursor_N_tuples-4.html}.

The datatypes of the results tuple is specified by a template given to
session_sql_query/4,5. See section~\ref{data-templates} for details on the
templates.  

\subsubsection
{\biptxtref{session_sql_query(+Session, +Template, +SQL, -Cursor)}{session_sql_query/4}{../bips/lib/dbi/session_sql_query-4.html}}
\label{session-sql-query/5}
%--------------------------------------------------------------------

This executes SQL and creates the handle Cursor for the SQL query. Template
specifies the datatypes of the results tuples.

\subsubsection
{\biptxtref{cursor_next_tuple(+Cursor, -Tuple)}{cursor_next_tuple/2}{../bips/lib/dbi/cursor_next_tuple-2.html}}
\label{cursor-next-tuple/2}
%--------------------------------------------------------------------

A single tuple is retrieved from the database.  Calling this
predicate again with the same cursor will retrieve further tuples
Any NULL values are returned as uninstantiated variables.

Once all the tuples have been retrieved this predicate fails.

If Tuple does not unify with the retrieved tuple, the predicate fails.

\begin{verbatim}
check_overdraft_limit(Session, Account) :-
    L = ["select count(id) from accounts \
        where     id = ",Account," and balance < overdraft"],
    concat_string(L,SQL),
    session_sql_query(Session,c(0),SQL,OverdraftCheck),
    cursor_next_tuple(OverdraftCheck,c(Count)),
    Count = 0.
\end{verbatim}
In this example a query is built to verify that the balance of an
account is not less than its overdraft facility. All comparisons
are done within the database, and we are just interested in checking
that no rows match the 'where' clause.

For this kind of application one would not normally use
concat_string/2. SQL placeholders would be used instead. See
\bipref{session_sql_prepare_query/5}{../bips/lib/dbi/session_sql_prepare_query-5.html}.


\subsubsection
{\biptxtref{cursor_all_tuples(+Cursor, -TupleList)}{cursor_all_tuples/2}{../bips/lib/dbi/cursor_all_tuples-2.html}}
\label{cursor-all-tuples/2}
%--------------------------------------------------------------------

The SQL query represented by the cursor is executed and all
the matching tuples are collected in TupleList.

This could be defined as:

\begin{verbatim}
cursor_all_tuples( Cursor, Tuples ) :-
    ( cursor_next_tuple(Cursor, T) ->
        Tuples = [T | Ts],
        cursor_all_tuples(Cursor, Ts)
    ;
        Tuples = []
    ).
\end{verbatim}

\subsubsection
{\biptxtref{cursor_N_tuples(+Cursor, -Retrieved, -TupleList, -RestTupleList)}{cursor_N_tuples/4}{../bips/lib/dbi/cursor_N_tuples-4.html}}
\label{cursor-N-tuples/4}
%--------------------------------------------------------------------

If the underlying DB supports the retrieving mutule tuples in one go, then
a buffer full of tuples matching the query is retrieved, otherwise all the
remaining tuples are retrieved.

TupleList and RestTupleList form a difference list containing these
tuples. Retrieved is the number of tuples retrieved.


%--------------------------------------------------------------------
\subsection{Parametrised Database Queries}
%--------------------------------------------------------------------

\begin{sloppypar}
The library allow SQL queries to be prepared and parameterised, if prepared
SQL statements are supported by the underlying database. Templates are needed
for specifying the datatypes of the parameters (as with
\bipref{session_sql_prepare/4}{../bips/lib/dbi/session_sql_prepare-4.html}),
and for the results tuples (as with \bipref{session_sql_query/4}{../bips/lib/dbi/session_sql_query-4.html}. An SQL query
is prepared by
\bipref{session_sql_prepare_query/5}{../bips/lib/dbi/session_sql_prepare_query-5.html},
it then needs to be executed by
\bipref{cursor_next_execute/2}{../bips/lib/dbi/cursor_next_execute-2.html} or
\bipref{cursor_next_execute/3}{../bips/lib/dbi/cursor_next_execute-3.html}
(cursor_next_execute/3 allows the specification of options for the cursor),
and the results can then be retrieved by \bipref{cursor_next_tuple/2}{../bips/lib/dbi/cursor_next_tuple-2.html},
\bipref{cursor_all_tuples/2}{../bips/lib/dbi/cursor_all_tuples-2.html} and
\bipref{cursor_N_tuples/4}{../bips/lib/dbi/cursor_N_tuples-4.html}. After
executing cursor_next_execute/2, it can be executed again with a new tuple
of parameter values. Any unretrieved results from the previous execute are
discarded. Note that this is non-logical: the discarded results are not
recovered on backtracking.
\end{sloppypar}

\subsubsection
{\biptxtref{session_sql_prepare_query(+Session, +ParamT, +QueryT, +SQL,-Cursor)}{session_sql_prepare_query/5}{../bips/lib/dbi/session_sql_prepare_query-5.html}}
\label{session-prepare-sql-query/5}
%--------------------------------------------------------------------

This creates Cursor, which is a handle to the prepared SQL query.

By changing the placeholders one gets a fresh query and can start extracting
tuples again.


In this example a generic query is built to check whether an account is
overdrawn, and a cursor for this query is created.
\begin{verbatim}
make_check_overdraft_limit(Session, Cursor) :-
    SQL = "select count(id) from accounts where ID = ? \
               and balance < overdraft",
    session_sql_prepare_query(Session,a(0),c(0),SQL,1,Cursor).

check_overdraft_limit(Check,Account) :-
    cursor_next_execute(Check,a(Account)),
    cursor_next_tuple(Check,c(Count)),
    Count = 0.
\end{verbatim}
The check_overdraft_limit/2 predicate uses the cursor to check an account.
This cursor can be re-used to check any number of accounts.


\index{database, interface to|)}
\index{MySQL, interface to|)}
%HEVEA\cutend
