% BEGIN LICENSE BLOCK
% Version: CMPL 1.1
%
% The contents of this file are subject to the Cisco-style Mozilla Public
% License Version 1.1 (the "License"); you may not use this file except
% in compliance with the License.  You may obtain a copy of the License
% at www.eclipse-clp.org/license.
% 
% Software distributed under the License is distributed on an "AS IS"
% basis, WITHOUT WARRANTY OF ANY KIND, either express or implied.  See
% the License for the specific language governing rights and limitations
% under the License. 
% 
% The Original Code is  The ECLiPSe Constraint Logic Programming System. 
% The Initial Developer of the Original Code is  Cisco Systems, Inc. 
% Portions created by the Initial Developer are
% Copyright (C) 2006 Cisco Systems, Inc.  All Rights Reserved.
% 
% Contributor(s): 
% 
% END LICENSE BLOCK

% File		: knowbase-lst.tex
% Date		: March 1992
% Author	: Michael Dahmen (edit of automatic generated file)
% Modified by	: Luis Hermosilla, August 1992
% Project	: MegaLog-Sepia User Manual
% Content	: predicates of module Deductive Relations

\section{\eclipse KB}

The following predicates are exported by the module {\bf kb}.

%\biponelinesection{\eclipse KB}
\begin{description}
\item[+RelationName \hspace {1mm} $<$$-$$-$$-$ \hspace {1mm} +Expression]{Set evaluation of a deductive algebraic expression and deletion from a deductive relation.}
\index{$<---$/2}
\item[+RelationName \hspace {1mm} $<$+++ \hspace {1mm} +Expression]{Set evaluation of a deductive algebraic expression and insertion into a deductive relation.}
\index{$<+++$/2}
\item[+RelationName1 \hspace {1mm} $<$@@$>$ \hspace {1mm} +RelationName2]{Test for schema compatibility.}
\index{$<@@>$/2}
\item[?RelationName \hspace {1mm} $<$$-$$-$$>$ \hspace {1mm} ?Synonym]{Add, remove or query synonyms for relations.}
\index{$<-->$/2}
\item[?RelationName \hspace {1mm} $<$==$>$ \hspace {1mm} ?SpecList)]{Create or remove a relation. Query relation format.}
\index{$<==>$/2}
\item[expand +Expression]{Set evaluation of a deductive algebraic expression and display on standard output.}
\index{expand/2}
\item[?RelationName expand +Expression]{Set evaluation of a deductive algebraic expression and creation of a deductive target relation.}
\index{expand/2}
\item[isdr +Expression]{Set evaluation of a deductive algebraic expression and display on standard output.}
\index{isdr/2}
\item[?RelationName isdr +Expression]{Set evaluation of a deductive algebraic expression and creation of a deductive target relation.}
\index{isdr/2}
\item[closekb]{Close a knowledge base and remove all temporary relations. Fails if no knowledge base is open.}
\index{closekb/0}
\item[createkb(+KnowledgeBasePath)]{Create a new knowledge base or open a knowledge base if it already exists.}
\index{createkb/1}
\item[define(+FileName)]{Add a Prolog source file to the definitions file.}
\index{define/1}
\item[define_implicit(+Relation/+Arity)]{Make a relation transparent.}
\index{define_implicit/1}
\item[degree(+RelationName, ?Degree)]{Get the number of attributes in a relation.}
\index{degree/2}
\item[delete_clause(+Clause)]{Delete a variant clause from a deductive relation.}
\index{delete_clause/1}
\item[destroykb]{Destroy the currently open knowledge base.}
\index{destroykb/0}
\item[display_defs]{Display the definitions file.}
\index{display_defs/0}
\item[helpdrel(+RelationName)]{Print some info on {\it RelationName} to the standard output.}
\index{helpdrel/1}
\item[helpkb]{Print a list of all relations to the standard output.}
\index{helpkb/0}
\item[insert_clause(+Clause)]{Insert a clause into a deductive relation.}
\index{insert_clause/1}
\item[insert_clauses_from(+FileName)]{Insert a file of program clauses into the deductive database.}
\index{insert_clauses_from/1}
\item[openkb(+KnowledgeBasePath)]{Open a knowledge base.}
\index{openkb/1}
\item[retract_clause(?Clause)]{Retract a unifying clause from a deductive relation.}
\index{retract_clause/1}
\item[retrieve_clause(?Clause)]{Retrieve a unifying clause from a deductive relation.}
\index{retrieve_clause/1}
\item[update_defs]{Edit the definitions file.}
\index{update_defs/0}
\end{description}

